---
title: 'Module 5: Introduction to MLM'
author: "Phil Mike Jones"
date: '`r Sys.Date()`'
output:
  html_document: default
  pdf_document: default
---

# Setup

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath("~/gits/multi_level_modelling/"))
options(scipen = 5, digits = 2)
```

Load packages:
```{r packages, warning=FALSE, message=FALSE}
library("magrittr")
library("readr")
library("lme4")
```

Load Scottish School Leavers Survey (SSLS) data:
```{r read-ssls, warning=FALSE, cache=TRUE, message=FALSE}
ssls <- readr::read_csv("inst/extdata/5.1.txt")
head(ssls)
```

Tidy dataset and configure factors:
```{r tidy-ssls}
ssls$female <- factor(ssls$female, levels = 0:1, labels = c("male", "female"))
ssls$sclass <- factor(ssls$sclass, levels = 1:4,
                      labels = c("managerial", "intermediate", "working",
                                 "unclassified"))
ssls$schtype <- factor(ssls$schtype, levels = 0:1,
                       labels = c("independent", "state"))
ssls$schurban <- factor(ssls$schurban, levels = 0:1,
                        labels = c("town_rural", "urban"))
ssls$schdenom <- factor(ssls$schdenom, levels = 0:1,
                        labels = c("non-denom", "roman_catholic"))
```

Check head and tail of data frame:
```{r check-ssls}
head(ssls)
tail(ssls)
```

# P5.1: Comparing Groups using Multilevel Modelling
## P5.1.1 A multi--level model of attainment with school effects

Set up a null model using only school effects. The null model in general is:

$$
score_{ij} = \beta_{0} + u_{oj} + e_{ij}
$$

where $score_{ij}$ is the attainment of student $i$ in school $j$, $\beta_{o}$ is the overall mean across schools, $u_{oj}$ is the effect of school $j$ on attainment, and $e_{ij}$ is student $ij$'s error term or residual.
The null model therefore attempts to explain student ${ij}$'s attainment by taking the mean, adjusting based on the attainment of the school they attend, and including an error term.

```{r null-model}
m_null <- lmer(score ~ (1 | schoolid), data = ssls, REML = FALSE)
```

In setting up the null model no fixed effects are included and the intercept is included by default.
The `(1 | schoolid)` refers to the constant (`1`) and random level 2 explanatory variables (`schoolid`).
The constant is an integer greater than or equal to one, with the greater the value the more accurate the estimation of the log--likelihood at the cost of greater computational time.
`REML = FALSE` uses a maximum likelihood estimation rather than a restricted maximum likelihood estimation, the default.

```{r null-summary}
summary(m_null)
```

* Formula displays the call used for the model.
* Data provides summary statistics for the model.
* Random effects show the effects of random effect variables on the model, in this case only `schoolid`.
* Fixed effects shows the effect of fixed effect variables on the model, in this case just an intercept.

## Comparing across schools
From the fixed effects the mean attainment value across all schools is estimated as `r m_null@beta`.
The mean for school $j$ is therefore estimated as $`r m_null@beta` + \hat{u}_{oj}$, where $\hat{u}_{oj}$ is the school residual.
A $\hat{u}_{oj} > 0$ means the school has a higher attainment than average, although at this point these are only point estimates without confidence intervals so it's not possible if the differences are 'real'.

```{r extract-variances, include=FALSE}
m_null_var <- VarCorr(m_null) %>% as.data.frame(.)

m_null_btwsch_var <- m_null_var$vcov[m_null_var$grp == "schoolid"]
m_null_wthsch_var <- m_null_var$vcov[m_null_var$grp == "Residual"]
```

* The between--school variance (level 2) in attainment is estimated as `r m_null_btwsch_var` and is denoted by $\hat{\sigma}^2_{u0}$

* The within--school between--student (level 1) variance is `r m_null_wthsch_var` and is denoted by $\hat{\sigma}^2_{e}$

* The total variance is the sum of these two, i.e. $`r m_null_btwsch_var` + `r m_null_wthsch_var` = `r m_null_btwsch_var + m_null_wthsch_var`$.

* The variance partition coefficient (VPC) is $\frac{\hat{\sigma}^2_{u0}}{(\hat{\sigma}^2_{u0} + \hat{\sigma}^2_{e})}$ or $\frac{`r m_null_btwsch_var`}{`r m_null_btwsch_var + m_null_wthsch_var`}$ = $`r m_null_btwsch_var / (m_null_btwsch_var + m_null_wthsch_var)`$.
This indicates that `r m_null_btwsch_var / (m_null_btwsch_var + m_null_wthsch_var) * 100`\% of the variance in attainment is attributed to differences between schools.

To test if this figure is statistically significant we carry out a **likelihood ratio test** comparing the null multi--level model (that we have just calculated) with a null single--level model.
The null single--level model simply exludes the school (which is a random effect):

$$
score_{ij} = \beta_{0} + e_{ij}
$$

Where $score_{ij}$ is again the attainment score of pupil $i$ in school $j$, $\beta_{0}$ is the mean attainment score, and $e_{ij}$ is pupil $ij$'s error term or residual.

```{r}
m_null_single <- lm(score ~ 1, data = ssls)
summary(m_null_single)
```

The **likelihood ratio test** statistic is two times the difference in the log likelihood values for the two models.

```{r}
anova(m_null, m_null_single) %>% str()
```

<!-- u0 <- ranef(m_null, postVar = TRUE) -->

<!-- u0se <- sqrt(attr(u0[[1]], "postVar")[1, , ]) -->

<!-- str(u0[1]) -->

<!-- str(u0[[1]]) -->

<!-- head(attr(u0[[1]], "postVar")[1, , ]) -->

<!-- schoolid <- as.numeric(rownames(u0[[1]])) -->

<!-- u0tab <- cbind(schoolid, u0[[1]], u0se) -->

<!-- colnames(u0tab) <- c("schoolid","u0","u0se") -->

<!-- u0tab <- u0tab[order(u0tab$u0), ] -->

<!-- u0tab <- cbind(u0tab, c(1:dim(u0tab)[1])) -->

<!-- colnames(u0tab)[4] <- "u0rank" -->

<!-- u0tab <- u0tab[order(u0tab$schoolid), ] -->

<!-- u0tab[1:10, ] -->

<!-- plot(u0tab$u0rank, u0tab$u0, type = "n", xlab = "u_rank", ylab = "conditional modes of r.e. for school_id:_cons") -->

<!-- segments(u0tab$u0rank, u0tab$u0 - 1.96*u0tab$u0se, u0tab$u0rank, u0tab$u0 + 1.96*u0tab$u0se) -->

<!-- points(u0tab$u0rank, u0tab$u0, col = "blue") -->

<!-- abline(h = 0, col = "red") -->

# Copyright and acknowledgements

These files are my own notes based on course content copyright 2011 Camille Szmaragd and George Leckie, Centre for Multilevel Modelling. These notes are not intended in any way to imply this work is my own original work (and in fact in places they copy word--for--word the course content).
