---
title: 'Module 5: Introduction to MLM'
author: "Phil Mike Jones"
date: "1 August 2016"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath("~/gits/multi_level_modelling/"))
```

```{r packages}
library("readr")
library("lme4")
```

# P5.1: Comparing Groups using Multilevel Modelling

Copyright 2011 Camille Szmaragd and George Leckie, Centre for Multilevel
Modelling

Load Scottish School Leavers Survey (SSLS) data

```{r read-data}
ssls <- readr::read_csv("inst/extdata/5.1.txt")
```

mydata <- read.table(file = "5.1.txt", sep = ",", header = TRUE)

str(mydata)


# P5.1.1 A multilevel model of attainment with school effects

library(lme4)

nullmodel <- lmer(score ~ (1 | schoolid), data = mydata, REML = FALSE)

summary(nullmodel)


# P5.1.2 Examining school effects (residuals)

fit <- lm(score ~ 1, data = mydata)

summary(fit)

logLik(nullmodel)

logLik(fit)

u0 <- ranef(nullmodel, postVar = TRUE)

u0se <- sqrt(attr(u0[[1]], "postVar")[1, , ])

str(u0[1])

str(u0[[1]])

head(attr(u0[[1]], "postVar")[1, , ])

schoolid <- as.numeric(rownames(u0[[1]]))

u0tab <- cbind(schoolid, u0[[1]], u0se)

colnames(u0tab) <- c("schoolid","u0","u0se")

u0tab <- u0tab[order(u0tab$u0), ]

u0tab <- cbind(u0tab, c(1:dim(u0tab)[1]))

colnames(u0tab)[4] <- "u0rank"

u0tab <- u0tab[order(u0tab$schoolid), ]

u0tab[1:10, ]

plot(u0tab$u0rank, u0tab$u0, type = "n", xlab = "u_rank", ylab = "conditional modes of r.e. for school_id:_cons")

segments(u0tab$u0rank, u0tab$u0 - 1.96*u0tab$u0se, u0tab$u0rank, u0tab$u0 + 1.96*u0tab$u0se)

points(u0tab$u0rank, u0tab$u0, col = "blue")

abline(h = 0, col = "red")
